\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-04-04 Paper/chapters/implementation-strategies.tex
benchmark
bibliography no ul

\section{Strategies for SPN}

@@

Nowadays, most instructions set architectures support single-instruction,
multiple-data processing. The idea of such an SIMD system is to work on
multiple data stored in vectors at once to speed up calculations. For A64, two
types of vector processing are available:

\begin{enumerate}
    \item Advanced SIMD, known as NEON
    \item Scalable Vector Extension (SVE)
\end{enumerate}

We will take a look at NEON as this is the type of vector processing supported
by the Cortex-A73 processor.

\subsubsection{ARM Neon}

The register file of the NEON unit is made up of 32 quad-word (128-bit)
registers \texttt{V[0-31]}, each extending the standard 64-bit floating-point
registers \mbox{\texttt{D[0-31]}}. These registers are divided into equally
sized lanes on which the vector instructions operate. Valid ways to interpret
for example the register \texttt{V0} are:

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{Figures/V_register.pdf}
    \caption{Divisions of the V0 register}
\end{figure}

NEON instructions interpret their operands' layouts (i.e. lane count and width)
through the use of suffixes such as \texttt{.4S} or \texttt{.8H}. For example,
adding eight 16-bit halfwords from register \texttt{V1} and \texttt{V2}
together and storing the result in \texttt{V0} can be done as follows:

\begin{center}
    \texttt{ADD V0.8H, V1.8H, V2.8H}
\end{center}

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{Figures/vector_add.pdf}
    \caption{Addition of two vector registers}
\end{figure}

The plenitude of different processing instructions allow flexible ways to
further speed up algorithms having reached their optimizational limit on
non-SIMD platforms. \texttt{vperm}, a general term standing for \textit{vector
permute}, is a common instruction on SIMD machines. Called \texttt{TBL} on
NEON, it is used for parallel table lookups and arbitrary permutations. It takes
two inputs to perform a lanewise lookup:

\begin{enumerate}
    \item A register with lookup values
    \item Two or more registers containing data
\end{enumerate}

@@

\[
    \scriptsize
    \begin{array}{c|llllllll|llllllll}
        n & 15 & 14 & 13 & 12 & 11 & 10 & 9 & 8 & 7 & 6 & 5 & 4 & 3 & 2 & 1 & 0 \\
        \hline
        V_0 & B56 & B48 & B40 & B32 & B24 & B16 & B8  & B0 & A56 & A48 & A40 & A32 & A24 & A16 & A8  & A0 \\
        V_1 & B57 & B49 & B41 & B33 & B25 & B17 & B9  & B1 & A57 & A49 & A41 & A33 & A25 & A17 & A9  & A1 \\
        V_2 & B58 & B50 & B42 & B34 & B26 & B18 & B10 & B2 & A58 & A50 & A42 & A34 & A26 & A18 & A10 & A2 \\
        V_3 & B59 & B51 & B43 & B35 & B27 & B19 & B11 & B3 & A59 & A51 & A43 & A35 & A27 & A19 & A11 & A3 \\
        V_4 & B60 & B52 & B44 & B36 & B28 & B20 & B12 & B4 & A60 & A52 & A44 & A36 & A28 & A20 & A12 & A4 \\
        V_5 & B61 & B53 & B45 & B37 & B29 & B21 & B13 & B5 & A61 & A53 & A45 & A37 & A29 & A21 & A13 & A5 \\
        V_6 & B62 & B54 & B46 & B38 & B30 & B22 & B14 & B6 & A62 & A54 & A46 & A38 & A30 & A22 & A14 & A6 \\
        V_7 & B63 & B55 & B47 & B39 & B31 & B23 & B15 & B7 & A63 & A55 & A47 & A39 & A31 & A23 & A15 & A7
    \end{array}
\]

>>>>

\section{Strategies for \texttt{GIFT}}

@@

The plenitude of different processing instructions introduced by
NEON\ref{ss:neon} allow flexible ways to further speed up algorithms having
reached their optimizational limit on non-SIMD platforms. \texttt{vperm}, a
general term standing for \textit{vector permute}, is a common instruction on
SIMD machines. Called \texttt{TBL} on NEON, it is used for parallel table
lookups and arbitrary permutations. It takes two inputs to perform a lanewise
lookup:

\begin{enumerate}
    \item A register with lookup values
    \item One or more registers containing data
\end{enumerate}

@@

\scriptsize
\[
    \begin{array}{c|llllllll|llllllll}
        n & 15 & 14 & 13 & 12 & 11 & 10 & 9 & 8 & 7 & 6 & 5 & 4 & 3 & 2 & 1 & 0 \\
        \hline
        V_0 & B56 & B48 & B40 & B32 & B24 & B16 & B8  & B0 & A56 & A48 & A40 & A32 & A24 & A16 & A8  & A0 \\
        V_1 & B57 & B49 & B41 & B33 & B25 & B17 & B9  & B1 & A57 & A49 & A41 & A33 & A25 & A17 & A9  & A1 \\
        V_2 & B58 & B50 & B42 & B34 & B26 & B18 & B10 & B2 & A58 & A50 & A42 & A34 & A26 & A18 & A10 & A2 \\
        V_3 & B59 & B51 & B43 & B35 & B27 & B19 & B11 & B3 & A59 & A51 & A43 & A35 & A27 & A19 & A11 & A3 \\
        V_4 & B60 & B52 & B44 & B36 & B28 & B20 & B12 & B4 & A60 & A52 & A44 & A36 & A28 & A20 & A12 & A4 \\
        V_5 & B61 & B53 & B45 & B37 & B29 & B21 & B13 & B5 & A61 & A53 & A45 & A37 & A29 & A21 & A13 & A5 \\
        V_6 & B62 & B54 & B46 & B38 & B30 & B22 & B14 & B6 & A62 & A54 & A46 & A38 & A30 & A22 & A14 & A6 \\
        V_7 & B63 & B55 & B47 & B39 & B31 & B23 & B15 & B7 & A63 & A55 & A47 & A39 & A31 & A23 & A15 & A7
    \end{array}
\]
\normalsize

### 2023-04-04 Paper/chapters/implementation.tex
xor instead of or
bibliography no ul

We will provide and discuss implementations for the presented strategies in the
C programming language. Although directly writing Assembler code could result
in a small performance benefit, this generally increases the work necessary by
an order of magnitude for only limited results. This is why we use NEON
intrinsics.

\subsection{NEON intrinsics}

The header file \texttt{<arm\_neon.h>} provides ARM-specific data and function
definitions including vector data types and C functions for working with these
vectors. These functions are known as NEON intrinsics and give the programmer a
high-level interface to most NEON instructions. Major advantages of this
approach include the ease of development as the compiler takes over register
allocation and load/store operations as well as performance benefits through
compiler optimizations.

Standard vector data types have the format \texttt{uintnxm\_t} with lane width
$n$ in bits and and lane count $m$. Array types of the format
\texttt{uintnxmxc\_t}, $c\in\{2,3,4\}$ are also defined which are used in
operations requiring multiple parameters like \texttt{TBL} or pairwise
load/stores. Intrinsics start with \texttt{v}, optionally contain a \texttt{q}
after the operation name to indicate operation on a 128-bit register, and end
with the lane data format. Multiplying eight pairs of 16-bit numbers
\texttt{a,b} for example can be done via the following:

\begin{center}
    \texttt{uint16x8\_t result = vmulq\_u16(a, b);}
\end{center}

In this case, the compiler allocates vector registers for \texttt{a},
\texttt{b} and \texttt{result} and assembles the intrinsic to \texttt{MUL
Vr.8H, Va.8H, Vb.8H}. Necessary load and stores for the result and parameters
are also handled automatically. Of special interest to us are the following
intrinsics, each existing in different variants with different lane widths and
also array types: \\

\begin{table}[h!]
    \centering
    \footnotesize
    \begin{tabularx}{\textwidth}{ll|X}
        Intrinsic && Description \\
        \hline
        \texttt{uint8x16\_t} & \texttt{vreinterpretq\_u8\_u64(uint64x2\_t)} & Explicit casting \\
        \hline
        \texttt{uint64\_t} & \texttt{vgetq\_lane\_u64(void)} & Extract a single lane \\
        \hline
        \texttt{void} & \texttt{vsetq\_lane\_u64(uint64\_t)} & Insert a single lane \\
        \hline
        \texttt{uint64x2\_t} & \texttt{vdupq\_n\_u64(uint64\_t)} & Initialize all lanes to same value \\
        \hline
        \texttt{void} & \texttt{vst1q\_u64(uint64\_t*, uint64x2\_t)} & Store from register to memory \\
        \hline
        \texttt{uint64x2\_t} & \texttt{vld1q\_u64(uint64\_t*, uint64x2\_t)} & Load from memory to register \\
        \hline
        \texttt{uint8x16\_t} & \texttt{veorq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise XOR \\
        \hline
        \texttt{uint8x16\_t} & \texttt{vandq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise AND \\
        \hline
        \texttt{uint8x16\_t} & \texttt{vorrq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise OR \\
        \hline
        \texttt{uint8x16\_t} & \texttt{vmvnq\_u8(uint8x16\_t)} & bitwise NOT \\
        \hline
        \texttt{uint8x16\_t} & \texttt{vqtbl2q\_u8(uint8x16\_t, uint8x16\_t)} & permutation (\texttt{TBL}) \\
    \end{tabularx}
\end{table}

The implementations closely follow the aforementioned strategies;
instruction-level optimization is left to the compiler.

>>>>

Implementations in the C programming language for the presented strategies can
be found in Appendix \ref{app:cimpl}. Although directly writing Assembler code could
result in a small performance benefit, this generally increases the work
necessary by an order of magnitude for only limited results. Instruction-level
optimization and in particular register allocation is left to the compiler.

### 2023-04-04 Paper/chapters/introduction.tex
bibliography no ul

Round key extraction differs for \texttt{GIFT-64} and \texttt{GIFT-128}. Let
$K=k7||k6||\dots||k0$ denote the $128$-bit key state.

\paragraph{GIFT-64}. We extract two 16-bit words $U||V=k_1||k_0$ from the key
state. $u_i$ and $v_i$ are XORed to $r_{4i+1}$ and $r_{4i}$ of the round key
$R$ respectively.

\paragraph{GIFT-128}. We extract two 32-bit words $U||V=k_5||k4||k1||k_0$ from
the key state. $u_i$ and $v_i$ are XORed to $r_{4i+2}$ and $b_{4i+1}$ of the
round key $R$ respectively.

@@

With small devices, embedded processors and ASICs becoming ever more ubiquitous
and essential in areas like medicine or automotive design, the need for ...

>>>>

Round key extraction differs for \texttt{GIFT-64} and \texttt{GIFT-128}. Let
$K=k_7||k_6||\dots||k_0$ denote the $128$-bit key state.

\paragraph{GIFT-64}. We extract two 16-bit words $U||V=k_1||k_0$ from the key
state. These are then added to the 64-bit round key: $R_{4i+1}\leftarrow
u_i,R_{4i}\leftarrow v_i$.

\paragraph{GIFT-128}. We extract two 32-bit words $U||V=k_5||k_4||k_1||k_0$ from
the key state. These are then added to the 128-bit round key: $R_{4i+2}\leftarrow
u_i,R_{4i+1}\leftarrow v_i$.

@@

With small devices and embedded processors becoming ever more ubiquitous and
essential in areas like consumer electronics or industrial and IoT
applications, the need for low-power, high-performance microprocessors has
increased steadily. With more than 250 billion chips shipped, semiconductors
designed by ARM power 95\% of mobile devices and have found a great many
applications due to their high performance and low power consumption. The
ODROID-N2+\cite{odroidn2} development board we are using is based on the big.LITTLE
architecture and is powered by a quad-core ARM Cortex-A73 processor and a
weaker dual-core ARM Cortex-A53 for power efficiency. Both these processors are
part of the eight generation of ARM designs known as ARMv8\cite{armv8:2013}.

ARMv8 defines three architecture profiles for different use cases as well as
dynamic execution states with corresponding instruction sets. This work will
focus on the A profile running in the AArch64 state utilizing the A64
instruction set with NEON and crypto extensions.

\begin{table}[h!]
    \centering
    \begin{tabularx}{\textwidth}{lX}
        \toprule
        Profile & Description \\
        \midrule
        Application (A) & Traditional use with virtual memory and privilege level support \\
        Real-time (R) & Real-time, low-latency, deterministic embedded systems \\
        Microcontroller (M) & Very low-power, fast-interrupt embedded systems \\
        \bottomrule
    \end{tabularx}
    \caption{ARMv8 profiles}
\end{table}

\begin{table}[h!]
    \centering
    \begin{tabularx}{\textwidth}{llX}
        \toprule
        Execution state & Usage & Instruction sets \\
        \midrule
        AArch32 & 32-bit compatibility & A32/T32 \\
        AArch64 & 64-bit & A64 \\
        \bottomrule
    \end{tabularx}
    \caption{ARMv8 execution states}
\end{table}

\subsection{General architecture}

ARMv8 is a RISC architecture employing simple data processing instructions
operating only on registers as well as dedicated load/store instructions to
transfer data from register to memory and back. This enables faster execution
of individual instructions, a simplier pipeline design, predictable instruction
timings and fewer addressing modes.

The A64 instruction set defines 31 64-bit general-purpose registers
\texttt{X0-X30} which can also be accessed as 32-bit registers \texttt{W0-W30}.
Values are loaded from and stored to memory using \texttt{LDR}/\texttt{STR}.
Data processing instructions generally use explicit output registers instead of
overwriting the first input register.

\begin{table}[h!]
    \centering
    \small
    \begin{tabularx}{\textwidth}{llX}
        \toprule
        Addressing mode & Example & Description \\
        \midrule
        Base register & \texttt{LDR W0, [X1]} & \texttt{W0 = *(X1);} \\
        Offset & \texttt{LDR W0, [X1, \#12]} & \texttt{W0 = *(X1 + 12);} \\
        Pre-indexing & \texttt{LDR W0, [X1, \#12]!} & \texttt{X1 += 12; W0 = *(X1)} \\
        Post-indexing & \texttt{LDR W0, [X1], \#12} & \texttt{W0 = *(X1); X1 += 12} \\
        \bottomrule
    \end{tabularx}
    \caption{AArch64 addressing modes}
\end{table}

\subsection{NEON}
\label{ss:neon}

ARMv8 supports single-instruction, multiple-data ARMv8 supports
single-instruction, multiple-data (SIMD) processing. These systems allow the
programmer to store multiple pieces of data in a vector and work on them in
parallel to speed up calculations. The A64 instruction set defines two possible
SIMD implementations.

\begin{enumerate}
    \item Advanced SIMD, known as NEON
    \item Scalable Vector Extension (SVE)
\end{enumerate}

We will take a look at NEON as this is the type of vector processing supported
by the Cortex-A73 processor.

The register file of the NEON unit is made up of 32 quad-word (128-bit)
registers \texttt{V[0-31]}, each extending the standard 64-bit floating-point
registers \mbox{\texttt{D[0-31]}}. These registers are divided into equally
sized lanes on which the vector instructions operate. Valid ways to interpret
for example the register \texttt{V0} are:

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{Figures/V_register.pdf}
    \caption{Divisions of the V0 register}
\end{figure}

NEON instructions interpret their operands' layouts (i.e. lane count and width)
through the use of suffixes such as \texttt{.4S} or \texttt{.8H}. For example,
adding eight 16-bit halfwords from register \texttt{V1} and \texttt{V2}
together and storing the result in \texttt{V0} can be done as follows:

\begin{center}
    \texttt{ADD V0.8H, V1.8H, V2.8H}
\end{center}

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{Figures/vector_add.pdf}
    \caption{Addition of two vector registers}
\end{figure}

\subsection{NEON Intrinsics}

The header file \texttt{<arm\_neon.h>} provides ARM-specific data and function
definitions including vector data types and C functions for working with these
vectors. These functions are known as NEON intrinsics \cite{neonintr:2022} and
give the programmer a high-level interface to most NEON instructions. Major
advantages of this approach include the ease of development as the compiler
takes over register allocation and load/store operations as well as performance
benefits through compiler optimizations.

Standard vector data types have the format \texttt{uintnxm\_t} with lane width
$n$ in bits and and lane count $m$. Array types of the format
\texttt{uintnxmxc\_t}, $c\in\{2,3,4\}$ are also defined which are used in
operations requiring multiple parameters like \texttt{TBL} or pairwise
load/stores. Intrinsics include the operation name and lane data format as well
as an optional \texttt{q} suffix to indicate operation on a 128-bit register.
Multiplying eight pairs of 16-bit numbers \texttt{a,b} for example can be done
via the following:

\begin{center}
    \texttt{uint16x8\_t result = vmulq\_u16(a, b);}
\end{center}

In this case, the compiler allocates vector registers for \texttt{a},
\texttt{b} and \texttt{result} and assembles the intrinsic to \texttt{MUL
Vr.8H, Va.8H, Vb.8H}. Necessary loads and stores for the result and parameters
are also handled automatically. Of special interest to us are the following
intrinsics, each existing in different variants with different lane widths and
also array types: \\

\begin{table}[h!]
    \centering
    \footnotesize
    \begin{tabularx}{\textwidth}{llX}
        \toprule
        Intrinsic && Description \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vreinterpretq\_u8\_u64(uint64x2\_t)} & Explicit casting \\
        \midrule
        \texttt{uint64\_t} & \texttt{vgetq\_lane\_u64(void)} & Extract a single lane \\
        \midrule
        \texttt{void} & \texttt{vsetq\_lane\_u64(uint64\_t)} & Insert a single lane \\
        \midrule
        \texttt{uint64x2\_t} & \texttt{vdupq\_n\_u64(uint64\_t)} & Initialize all lanes to same value \\
        \midrule
        \texttt{void} & \texttt{vst1q\_u64(uint64\_t*, uint64x2\_t)} & Store from register to memory \\
        \midrule
        \texttt{uint64x2\_t} & \texttt{vld1q\_u64(uint64\_t*, uint64x2\_t)} & Load from memory to register \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{veorq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise XOR \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vandq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise AND \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vorrq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise OR \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vmvnq\_u8(uint8x16\_t)} & bitwise NOT \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vqtbl2q\_u8(uint8x16\_t, uint8x16\_t)} & permutation (\texttt{TBL}) \\
        \bottomrule
    \end{tabularx}
    \caption{Common NEON intrinsics}
\end{table}
